 

### 2021-06-13 vertical-thermal-review.Rmd

Biophysical conditions can alter tree-water dynamics and growth patterns. Greater $VPD$, atmospheric demand and low RH can be more stressful for trees, than soil moisture across biomes-including mesic forests,  as the former limits surface conductance and evapotranspiration to a greater extent than soil moisture [@novickIncreasingImportanceAtmospheric2016; @corlettImpactsWarmingTropical2011; @ruehrWaterAvailabilityDominant2016]. Along with greater VPD, increase in temperature even under adequete water can cause greater tree hydraulic stress due to altered soil-water dynamics than milder drought, in temperate species. Recovery time from heat stress varies between species hydraulic traits [@ruehrWaterAvailabilityDominant2016; @parkwilliamsTemperaturePotentDriver2013]. Whereas, greater $RH$ can do the opposite, where it stifles evapotranspiration thereby increasing $T_{leaf}$ [Fig.3; @perezIncreasingHumidityThreatens2018; @tibbittsHumidityPlants1979]. Altered precipitation regimes causing drought is more detrimental to larger trees more than small trees, because limited soil-water cannot meet the evaporative demand via the long-distance water transport to the canopy [Norway spruce study, @bennettLargerTreesSuffer2015; @mcgregor_tree_nodate; @bartholomewSmallTropicalForest2020]. However, given adequate moisture to sustain high $g_s$, canopy leaves exposed to more wind and lower humidity should be more effective at cooling when $T_{air}$ exceeds physiological optima (Fig. `r fig_leaf_T`).

Tolerance to multiple biophysical and biological stressors can differ among tree size classes, which has shown to increase with tree height and light due to greater non-structural carbon (NSC) reserves, with smaller understory trees being more sensitive to stress [@niinemetsResponsesForestTrees2010a]. This can also alter tree growth patterns. Particularly higher temperatures has shown to primarily reduce tree growth by altering water availability, in mesic temperate and subtropical eucalytus forests [@bowmanWarmerWorldWill2014]. Evidence in tropical forests also shows that higher nighttime temperatures is associated with decelarating tree growth rates predominantly in smaller trees but also in taller trees, perhaps due to increase in nighttime metabolic respiration that could supresses growth [*Clark 2010; 2013; Feeley et al. 2007*]. .

@@

Phenologically in temperate deciduous and mesic forests, understory growth can be more sensitive to annual warming than canopy, showing earlier spring onset and later autumn senescense [Zohnner and Renner, 2019]. Increase in summer temperature is also associated with reduced forest carbon storage [@oishiWarmerTemperaturesReduce2018]. However, late spring, short-term heat waves can negatively impact partially-expanded canopy leaves more than understory saplings that are buffered, by inducing leaf shedding and decreasing photosynthesis and C uptake in reflushed leaves [@filewodImpactsSpringHeat2014].

@@

Feedback loops from loss of big trees through natural disturbance or anthropogenic (e.g.@senfCanopyMortalityHas2018a, 2020;  @mcdowellMechanismsPlantSurvival2008; Brienen et al. 2015; @bennettLargerTreesSuffer2015; @stovallTreeHeightExplains2019a] would increase canopy roughness, and reductions in canopy cover is associated with reduced canopy structural complexity [@senfMappingForestDisturbance2020]. Leading to altered biophysical gradient with local heating effects [@zellweger_forest_2020; @starkReframingTropicalSavannization2020], non-linear threshold responses, causing dramatic shifts from one forest state to another (e.g., transition from forest to savanna-like vegetation in tropical forest regions through "savannization"), with energy balance impacts and implications for forest-atmosphere interactions [@starkReframingTropicalSavannization2020].

In the face of global climate change, resiliency among forests are also observed, such as African rainforests [Amy Bennett's new paper] that are more resilient to warming and drought than other forests. Increasing temperature acclimation among tropical and boreal species [martijn's papers; Danielle Way's papers; Atkin's papers] and reversible recovery of canopy heat stress [Smith et al. 2020] are other mechanisms [for eg. *REF*] that will help forests cope with warming.

>>>>

Biophysical conditions can alter tree-water dynamics and growth patterns. Greater $VPD$, atmospheric demand and low RH can be more stressful for trees, than soil moisture across biomes-including mesic forests,  as the former limits surface conductance and evapotranspiration to a greater extent than soil moisture [@novickIncreasingImportanceAtmospheric2016; @corlettImpactsWarmingTropical2011; @ruehrWaterAvailabilityDominant2016]. Along with greater VPD, increase in temperature even under adequete water can cause greater tree hydraulic stress due to altered soil-water dynamics than milder drought, in temperate species [@ruehrWaterAvailabilityDominant2016; @parkwilliamsTemperaturePotentDriver2013]. Whereas, greater $RH$ can do the opposite, where it stifles evapotranspiration thereby increasing $T_{leaf}$ [Fig.3; @perezIncreasingHumidityThreatens2018; @tibbittsHumidityPlants1979]. Altered precipitation regimes causing drought is more detrimental to larger trees more than small trees, because limited soil-water cannot meet the evaporative demand via the long-distance water transport to the canopy [Norway spruce study, @bennettLargerTreesSuffer2015; @mcgregor_tree_nodate; @bartholomewSmallTropicalForest2020]. However, given adequate moisture to sustain high $g_s$, canopy leaves exposed to more wind and lower humidity should be more effective at cooling when $T_{air}$ exceeds physiological optima (Fig. `r fig_leaf_T`).

Tolerance to multiple biophysical and biological stressors can differ among tree size classes, which has shown to increase with tree height and light, and resultant non-structural carbon (NSC) reserves, with smaller understory trees being more sensitive to stress [@niinemetsResponsesForestTrees2010a]. Autotrophic respiration also increases with stand age that declines NPP, whereas GPP increases with associated morphological and anatomical complexity with age [PW West, 2019]. This difference can alter tree growth patterns in response to biophysical conditions. Particularly higher temperatures has shown to primarily reduce tree growth by altering water availability, in mesic temperate and subtropical eucalytus forests [@bowmanWarmerWorldWill2014]. Evidence in tropical forests also shows that higher nighttime temperatures is associated with decelarating tree growth rates predominantly in smaller trees but also in taller trees, perhaps due to increase in nighttime metabolic respiration that could supresses growth [*Clark 2010; 2013; Feeley et al. 2007*]. .

@@

Phenologically in temperate deciduous and mesic forests, understory growth can be more sensitive to annual warming than canopy, showing earlier spring onset and later autumn senescense [Zohnner and Renner, 2019].  However, late spring, short-term heat waves can negatively impact partially-expanded canopy leaves more than understory saplings that are buffered, by inducing leaf shedding and decreasing photosynthesis and C uptake in reflushed leaves [@filewodImpactsSpringHeat2014]. Increase in summer temperature is also associated with reduced forest carbon storage [@oishiWarmerTemperaturesReduce2018].

@@

Feedback loops from loss of big trees through natural or anthropogenic disturbance (e.g.@senfCanopyMortalityHas2018a, 2020;  @mcdowellMechanismsPlantSurvival2008; Brienen et al. 2015; @bennettLargerTreesSuffer2015; @stovallTreeHeightExplains2019a] would increase canopy roughness, and reductions in canopy cover is associated with reduced canopy structural complexity [@senfMappingForestDisturbance2020]. Leading to altered biophysical gradient with local heating effects [@zellweger_forest_2020; @starkReframingTropicalSavannization2020], non-linear threshold responses, causing dramatic shifts from one forest state to another (e.g., transition from forest to savanna-like vegetation in tropical forest regions through "savannization"), with energy balance impacts and implications for forest-atmosphere interactions [@starkReframingTropicalSavannization2020].

In the face of global climate change, resiliency among forests also stand out, such as African rainforests [Amy Bennett's new paper] that are more resilient to warming and drought than other forests. Increasing temperature acclimation among tropical and boreal species [martijn's papers; Danielle Way's papers; Atkin's papers] and reversible recovery of canopy heat stress [Smith et al. 2020] are other mechanisms [for eg. *REF*] that will help forests cope with warming.
